A Decision Procedure for (Co)datatypes in SMT Solvers\thanks{In memoriam Morgan Deters 1979--2015}

### 2016-02-10 sister/sister.tex
more
clearer
more polish

\begin{abstract}
Datatypes and codatatypes are useful to represent finite and potentially
infinite objects. We describe a decision procedure
to reason about such types. The procedure has been
integrated in CVC4, a modern SMT (satisfiability modulo theories)
solver, which can be used both as a constraint solver and as an
automatic theorem prover. An evaluation based on formalizations developed
in the Isabelle proof assistant demonstrates the potential of the
procedure.
\end{abstract}

@@

In the past decade, satisfiability modulo theories (SMT) solvers
\cite{nieuwenhuis-et-al-2006}
% \cite{ganzinger-et-al-2004}
have emerged as one of the most powerful ways to
prove theorems in first-order logic. These solvers are also useful for model
finding, or constraint solving: Given a set of constraints, they can be used
to find a variable assignment that satisfy them.

A great benefit of the SMT approach is that it provides a flexible framework
for composing optimized decision procedures and other solvers for various
theories, including equality, linear arithmetic on $\mathbb{Z}$ and
$\mathbb{R}$, and first-order quantifiers ($\forall$,~$\exists$). In this
context, two recurrent scientific questions are:
\begin{itemize}
\item Which theories would be useful to applications?
\item Which theories could be implemented efficiently in \\ %%% TYPESETTING: hack
SMT solvers?
\end{itemize}

This paper suggests the following partial answer:\ freely generated algebraic
and coalgebraic datatypes---also called \emph{datatypes} and
\emph{codatatypes}, respectively.
Datatypes are ubiquitous in functional programs and formal specifications.
They are useful to represent finite data structures in computer
science but also arise in formalized mathematics.
And to represent infinite objects,
a natural choice is to turn to codatatypes,
their non-well-founded dual.
%
Despite their reputation for esotericism, codatatypes have a
role to play in computer science. For example,
Leroy's verified C compiler % Comp\-Cert
\shortcite{leroy-2009} %, the verified Java compiler Jinja\-Threads
%\cite{lochbihler-2010-jinja},
and Lochbihler's formalized Java memory model
\shortcite{lochbihler-2014-jmm} both depend on codatatypes to express infinite
processes.

@@

Datatypes and codatatypes share many properties, so it makes sense
to consider them together. There are, however, at least three important
differences.
%
First, \emph{codatatypes need not be well-founded.}
For example, the type
$$\vthinspace\keyw{codata}~\;\ty{stream}_{\vthinspace\tau} \,=\allowbreak\, \const{SCons}(\tau,\allowbreak\: \ty{stream}_{\vthinspace\tau})\vthinspace$$
of infinite sequences or streams over an element type $\tau$ is
allowed, even though it has no base case.
Second, \emph{a uniqueness rule takes the place of the acyclicity rule of datatypes.}
Cyclic constraints such as
$x \teq \const{S}(x)$ %, where $\const{C}$ is a constructor,
are unsatisfiable for datatypes,
thanks to an acyclicity rule,
but satisfiable for codatatypes. 
For the latter, a uniqueness rule ensures that two values
having the same infinite expansion must be equal; from $x
\teq \const{S}(x)$ and
$y \teq \const{S}(y)$, it deduces $x \teq y$.
These two rules cannot be finitely axiomatized, so they naturally belong in
a decision procedure.
%    * in particular, acyclicity and uniqueness are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
%(For the other (co)datatype properties---the injectivity, distinctness, and
%exhaustiveness of constructors and the selector laws---it is
%widely recognized that decision procedures can be more efficient than
%axiomatizations.)
%axiom.
%
Third, \emph{it must be possible to express cyclic values as closed terms and
to enumerate them.} %This is necessary when generating models.
The \hbox{$\mu$-binder} notation associates a name with
a subterm, making it possible to represent cyclic values.
For example,
the $\mu$-term $\const{SCons}(1,\: \MU s.\; \const{SCons}(0,\: \const{SCons}(9,\: s)))$
stands for the %lasso-shaped
sequence $1, 0, 9, 0, 9, 0, 9, \dotsc$.

@@

Our setting is a monomorphic (many-sorted) first-order logic.
We fix a signature %$\Sig = (\Types, \Funcs)$
consisting of a set of types $\Types$ and a
set of function symbols $\Funcs$\negvthinspace.
The types are partitioned into
$\Types = \Data \mathrel{\uplus} \Codata \mathrel{\uplus} \Nondata$, where $\Data$ are the
\emph{datatypes}, $\Codata$ are the \emph{codatatypes}, and $\Nondata$ are the remaining
\emph{ordinary types}.
% (which can be interpreted or not).
The function symbols are
partitioned into $\Funcs = \Ctr \mathrel{\uplus} \Sel$, where $\Ctr$ are the
\emph{constructors} and $\Sel$ are the \emph{selectors}. There is no need to
consider further function symbols because they can be abstracted away as
variables when combining theories.
%Exceptionally, it is convenient to use numeric constants ($0$, $1$, \ldots)\ 
%in
%examples.
%$\Sig$-terms are standard first-order terms
%over $\Sig$,
%without $\mu$-binders.
%(Section~\ref{sec:implementation-as-a-theory-solver-in-cvc4}).

@@

Datatypes and codatatypes share many basic properties. All properties
below are implicitly universally quantified and range over all $i$, $j$,
$\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\!\begin{aligned}[t]
\text{Distinctness:}\enskip
  & %\forall \bar x, \bar y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \quad\text{if $j \not= \jPrime$}}
  \\[-0.5\jot]
\text{Injectivity:}\enskip
  & %\forall x_1,\dotsc,x_{n_{i\negvthinspace j}} y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(x_1,\dotsc,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(y_1,\dotsc,y_{n_{i\negvthinspace j}}) \medrightarrow x_k \teq y_k}
  \\[-0.5\jot]
\text{Exhaustiveness:}\enskip
  & \smash{\const{d}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \const{d}_{im_i}(x)}
  \\[-0.5\jot]
\text{Selection:}\enskip
  & \smash{\const{s}_{i\negvthinspace j}^{\,k}(\const{C}_{i\negvthinspace j}(x_1,\dotsc,x_{n_{i\negvthinspace j}})) \teq x_k}
\end{aligned}
\]
%
%In the algebraic jargon, exhaustiveness helps ensure that ``no
%junk'' exists, whereas distinctness and injectivity guarantee that ``no
%confusion'' can arise.
%The result of selectors applied to the wrong
%constructor is left completely unspecified.
%
%
%  * how to deal with "wrong" selectors, e.g.
%        hd(nil1) = hd(nil2)?
%      * leave them unspecified; hence hd(nil1) = hd(nil2) in some models,
%        and not in other models
%
%
Datatypes are additionally characterized by an induction principle.
For the natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction prohibits models that contain infinite values.
% $\const{S}(\const{S}(\ldots))$.
%
For codatatypes, the dual notion is called coinduction:
Two values that yield the same observations must be equal, 
where the observations are made through selectors and discriminators.
Codatatypes are also guaranteed to contain all values corresponding to 
infinite variable-free constructor terms.

@@

Given a fixed signature, %~$\Sig$,
the decision procedure for the universal theory of (co)data\-types %$\thD$
determines the
%$\thD$-
satisfiability of finite sets $\Ec$ of literals:\ equalities
and disequalities between first-order terms, %$\Sig$-terms,
whose variables are interpreted existentially.
The decision procedure is formulated as an %tableau-like
abstract calculus.
%Following the tradition of refutational theorem proving,
Proving a universal quantifier-free conjecture is reduced to showing that
its negation is unsatisfiable. 
%The presentation is
%inspired by Barrett {\em et al.} \shortcite{barrett-et-al-2007} but at
%a higher level, using unoriented equations instead of oriented ones.

@@

Our calculus for the theory of (co)datatypes
consists of derivation rules.
A derivation rule can be applied to $\Ec$ if %all
the specified
premises are met.
The conclusion either specifies equalities to be added to $\Ec$
or is $\bot$ (contradiction).
% (in which case we call an application of it \emph{nonterminal})
% (in which case we call an application of it \emph{terminal}).
One of the rules has multiple conclusions, % separated by $\ror$,
denoting branching.
%
An application of a rule is \emph{redundant} if one of its non-$\bot$
conclusions leaves $\Ec$ unchanged.
A \emph{derivation tree} is a %finite
tree whose nodes are finite sets of
equalities, such that child nodes are obtained by a nonredundant application of a
derivation rule to the parent. A derivation tree is \emph{closed} if all of
its leaf nodes are $\bot$. A node is \emph{saturated} if no nonredundant
instance of a rule can be applied to it.

@@

In conjunction with \rn{Refl}, \rn{Sym}, and \rn{Trans}, the \rn{Cong} rule computes the congruence (upward) closure,
whereas the \rn{Inject} and \rn{Clash} rules %together
compute the unification (downward) closure.
For unification, %additional
equalities are inferred based on the injectivity of constructors by \rn{Inject},
and failures to unify equated terms are recognized by \rn{Clash}.
\rn{Conflict} recognizes when an equality and its negation both occur in $\Ec$, in which case $\Ec$ has no model.

@@

\begin{figure}[b!]
%\vspace*{+6pt} %% TYPESETTING HACK
%\small
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t : \delta
  \betweenantes
  t \in \tEc
  \betweenantes
  \Ctr^\delta = \{ \const{C}_1, \dotsc, \const{C}_m \}
\\
  \bigl( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\delta \bigr)
  \text{ or }
  \bigl( \delta \in \Data \text{ and } %\left|
  \delta
  %\right|
  \text{ is finite} \bigr)
}{
  {\mathop{\big|\kern-.25ex\big|}\vthinspace}_{i=1}^n\, \Ec := \Ec,\: \DISC{i} %\ROR
 % \ROR \cdots \ROR \Ec := \Ec,\: \DISC{m}
}
\)
\rn{Split}
\\[5\jot]
\(
\inferrule{
  t, u : \delta \in \Codata
  \betweenantes
  t, u \in \tEc
  \betweenantes
  \delta \text{ is a singleton}
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Single}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

>>>>

\begin{abstract}
Datatypes and codatatypes are useful to represent finite and potentially
infinite objects. We describe a decision procedure
to reason about such types. The procedure has been
integrated into CVC4, a modern SMT (satisfiability modulo theories)
solver, which can be used both as a constraint solver and as an
automatic theorem prover. An evaluation based on formalizations developed
in the Isabelle proof assistant shows the potential of the
procedure.
\end{abstract}

@@

In the past decade, satisfiability modulo theories (SMT) solvers
\cite{nieuwenhuis-et-al-2006}
% \cite{ganzinger-et-al-2004}
have emerged as one of the most powerful ways to
prove theorems in first-order logic. These solvers are also useful for model
finding, or constraint solving: Given some constraints, they can be used
to find a variable assignment that satisfy them.

A great benefit of the SMT approach is that it provides a flexible framework
for composing dedicated decision procedures and other solvers for various
theories, including equality, linear arithmetic on $\mathbb{Z}$ and
$\mathbb{R}$, and first-order quantifiers ($\forall$,~$\exists$). In this
context, two recurrent scientific questions are:
\begin{itemize}
\item Which theories would be useful to applications?
\item Which theories could be implemented efficiently in \\ %%% TYPESETTING: hack
SMT solvers?
\end{itemize}

This paper suggests the following partial answer:\ freely generated algebraic
and coalgebraic datatypes---also called \emph{datatypes} and
\emph{codatatypes}, respectively.
Datatypes are ubiquitous in functional programming and formal specifications.
They are useful to represent finite data structures in computer
science but also arise in formalized mathematics.
And to represent infinite objects,
a natural choice is to turn to codatatypes,
their non-well-founded dual.
%
Despite their reputation for esotericism, codatatypes have a
role to play in computer science. For example,
Leroy's verified C compiler % Comp\-Cert
\shortcite{leroy-2009} %, the verified Java compiler Jinja\-Threads
%\cite{lochbihler-2010-jinja},
and Lochbihler's formalized Java memory model
\shortcite{lochbihler-2014-jmm} both depend on codatatypes to express infinite
processes.

@@

Datatypes and codatatypes share many properties, so it makes sense
to consider them together. There are, however, at least three important
differences.
%
First, \emph{codatatypes need not be well-founded.}
For example, the type
$$\vthinspace\keyw{codata}~\;\ty{stream}_{\vthinspace\tau} \,=\allowbreak\, \const{SCons}(\tau,\allowbreak\: \ty{stream}_{\vthinspace\tau})\vthinspace$$
of infinite sequences or streams over an element type $\tau$ is
allowed, even though it has no base case.
Second, \emph{a uniqueness rule takes the place of the acyclicity rule of datatypes.}
Cyclic constraints such as
$x \teq \const{S}(x)$ %, where $\const{C}$ is a constructor,
are unsatisfiable for datatypes,
thanks to an acyclicity rule,
but satisfiable for codatatypes. 
For the latter, a uniqueness rule ensures that two values
having the same infinite expansion are equal; from $x
\teq \const{S}(x)$ and
$y \teq \const{S}(y)$, it deduces $x \teq y$.
These two rules cannot be finitely axiomatized, so they naturally belong in
a decision procedure.
%    * in particular, acyclicity and uniqueness are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
%(For the other (co)datatype properties---the injectivity, distinctness, and
%exhaustiveness of constructors and the selector laws---it is
%widely recognized that decision procedures can be more efficient than
%axiomatizations.)
%axiom.
%
Third, \emph{it must be possible to express cyclic values as closed terms and
to enumerate them.} %This is necessary when generating models.
The \hbox{$\mu$-binder} notation associates a name with
a subterm, making it possible to represent cyclic values.
For example,
the $\mu$-term $\const{SCons}(1,\: \MU s.\; \const{SCons}(0,\: \const{SCons}(9,\: s)))$
stands for the %lasso-shaped
sequence $1, 0, 9, 0, 9, 0, 9, \dotsc$.

@@

Our setting is a monomorphic (many-sorted) first-order logic.
We fix a signature %$\Sig = (\Types, \Funcs)$
consisting of a set of types and a
set of function symbols.
The types are partitioned into
the \emph{datatypes} $\Data$,
the \emph{codatatypes} $\Codata$,
and the are the remaining \emph{ordinary types} $\Nondata$.
% (which can be interpreted or not).
The function symbols are
partitioned into the
\emph{constructors} $\Ctr$ and the \emph{selectors} $\Sel$.
There is no need to consider further function symbols because they can be
abstracted away as variables when combining theories.
%Exceptionally, it is convenient to use numeric constants ($0$, $1$, \ldots)\ 
%in
%examples.
%$\Sig$-terms are standard first-order terms
%over $\Sig$,
%without $\mu$-binders.
%(Section~\ref{sec:implementation-as-a-theory-solver-in-cvc4}).

@@

Datatypes and codatatypes share many basic properties. All properties
below are implicitly universally quantified and range over all $i$, $j$,
$\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\!\begin{aligned}[t]
\text{Distinctness:}\enskip
  & %\forall \bar x, \bar y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \quad\text{if $j \not= \jPrime$}}
  \\[-0.5\jot]
\text{Injectivity:}\enskip
  & %\forall x_1,\dotsc,x_{n_{i\negvthinspace j}} y.\;\,
    \smash{\const{C}_{i\negvthinspace j}(x_1,\dotsc,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(y_1,\dotsc,y_{n_{i\negvthinspace j}}) \medrightarrow x_k \teq y_k}
  \\[-0.5\jot]
\text{Exhaustiveness:}\enskip
  & \smash{\const{d}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \const{d}_{im_i}(x)}
  \\[-0.5\jot]
\text{Selection:}\enskip
  & \smash{\const{s}_{i\negvthinspace j}^{\,k}(\const{C}_{i\negvthinspace j}(x_1,\dotsc,x_{n_{i\negvthinspace j}})) \teq x_k}
\end{aligned}
\]
%
%In the algebraic jargon, exhaustiveness helps ensure that ``no
%junk'' exists, whereas distinctness and injectivity guarantee that ``no
%confusion'' can arise.
%The result of selectors applied to the wrong
%constructor is left completely unspecified.
%
%
%  * how to deal with "wrong" selectors, e.g.
%        hd(nil1) = hd(nil2)?
%      * leave them unspecified; hence hd(nil1) = hd(nil2) in some models,
%        and not in other models
%
%
Datatypes are additionally characterized by an induction principle.
For the natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction prohibits infinite values $\const{S}(\const{S}(\ldots))$.
%
For codatatypes, the dual notion is called coinduction:
Two values that yield the same observations must be equal, 
where the observations are made through selectors and discriminators.
Codatatypes are also guaranteed to contain all values corresponding to 
infinite variable-free constructor terms.

@@

Given a fixed signature, %~$\Sig$,
the decision procedure for the universal theory of (co)data\-types %$\thD$
determines the
%$\thD$-
satisfiability of finite sets $\Ec$ of constraints:\ equalities ($\teq$)
and disequalities ($\not\teq$) between first-order terms, %$\Sig$-terms,
whose variables are interpreted existentially.
The decision procedure is formulated as an %tableau-like
abstract calculus.
%Following the tradition of refutational theorem proving,
Proving a universal quantifier-free conjecture is reduced to showing that
its negation is unsatisfiable. 
%The presentation is
%inspired by Barrett {\em et al.} \shortcite{barrett-et-al-2007} but at
%a higher level, using unoriented equations instead of oriented ones.

@@

Our calculus for the theory of (co)datatypes
consists of derivation rules.
A derivation rule can be applied to $\Ec$ if %all
the specified
premises are met.
The conclusion either specifies equalities to be added to $\Ec$
or is $\bot$ (contradiction).
% (in which case we call an application of it \emph{nonterminal})
% (in which case we call an application of it \emph{terminal}).
One of the rules has multiple conclusions separated by $\left|\kern-.1ex\right|$,
denoting branching.
%
An application of a rule is \emph{redundant} if one of its non-$\bot$
conclusions leaves $\Ec$ unchanged.
A \emph{derivation tree} is a %finite
tree whose nodes are finite sets of
equalities, such that child nodes are obtained by a nonredundant application of a
derivation rule to the parent. A derivation tree is \emph{closed} if all of
its leaf nodes are $\bot$. A node is \emph{saturated} if no nonredundant
instance of a rule can be applied to it.

@@

In conjunction with \rn{Refl}, \rn{Sym}, and \rn{Trans}, the \rn{Cong} rule computes the congruence (upward) closure,
whereas the \rn{Inject} and \rn{Clash} rules %together
compute the unification (downward) closure.
For unification, %additional
equalities are inferred based on the injectivity of constructors by \rn{Inject},
and failures to unify equated terms are recognized by \rn{Clash}.
\rn{Conflict} recognizes when an equality and its negation both occur in $\Ec$, in which case $\Ec$ is
unsatisfiable.

@@

\begin{figure}[b!]
%\vspace*{+6pt} %% TYPESETTING HACK
%\small
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t : \delta
  \betweenantes
  t \in \tEc
  \betweenantes
  \Ctr^\delta = \{ \const{C}_1, \dotsc, \const{C}_m \}
\\
  \bigl( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\delta \bigr)
  \text{ or }
  \bigl( \delta \in \Data \text{ and } %\left|
  \delta
  %\right|
  \text{ is finite} \bigr)
}{
  {\mathop{\big|\kern-.25ex\big|}\vthinspace}_{i=1}^n\, \Ec := \Ec,\: \DISC{i} %\ROR
 % \ROR \cdots \ROR \Ec := \Ec,\: \DISC{m}
}
\)
\rn{Split}
\\[6\jot]
\(
\inferrule{
  t, u : \delta \in \Codata
  \betweenantes
  t, u \in \tEc
  \betweenantes
  \delta \text{ is a singleton}
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Single}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}
